---
title: "ExploreCensus - MATH23C Final Project Highlights"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Visualizing Median Household Incomes

Upon cleaning up the dataset, we began our analysis by visualizing the median household incomes using a histogram. Overlaying the probability density graph for the normal distribution over the histogram, it became clear that median household incomes do not fall within this distribution.

```{r, echo=FALSE}

  clean <- read.csv("clean.csv")
## Histogram with Normal Distribution added
  # Display Histogram
  hist(clean$Income, breaks = "FD", probability = TRUE, col="light blue", xlab = "Income (USD)", main = "Histogram of Income Distribution")
  # To find the best fitting normal distribution, compute the mean and variance of the data
  mu <- mean(clean$Income)
  var <- var(clean$Income)
  sigma <- sd(clean$Income)     #estimates square root of the population variance
  curve(dnorm(x, mu, sigma), from = 0, to = 250000, add = TRUE, col = "red")
```

## Visualizng the Percentage of a Race in a Tract compared the Average Income of the Tract
We used scatterplots to visualize the percentage of a specific race in a tract with respect to the median income of the tract overall. In particular, we found that the slope for the Asian population was the most positive of all the scatterplots, reflecting broader trends amongst one of the fastest growing, highest-educated minority groups in the nation.

```{r, echo=FALSE}
clean <- read.csv("clean.csv")
if (!require(ggplot2)) install.packages("ggplot2"); library(ggplot2)
percentA <- clean$Asian / 100       # Asian
  percentB <- clean$Black / 100       # Black
  percentH <- clean$Hispanic / 100    # Hispanic
  percentN <- clean$Native / 100      # Native
  percentP <- clean$Pacific / 100     # Pacific
  percentW <- clean$White / 100       # White
  income <-  clean$Income
# ggplotting the data: fulfilling "Nicely labeled graphics using ggplot, with good use of color, line styles, etc., that tell a convincing story"
  df <- data.frame(
    percentA,      # Asian
    percentB,      # Black
    percentH,   # Hispanic
    percentN,    # Native
    percentP,  # Pacific
    percentW,       # White
    income
  )
  # Asian
  ggplot(df, aes(percentA, income)) + 
    geom_point(color = "red", size = 0.5) + 
    geom_smooth(method = "lm", color ="blue") +
    xlab("Percent of Asian People in Tract") +
    ylab("Average Income of Tract") + 
    ggtitle("Percent Asians vs. Average Income (in tract) ")
  # Black
  ggplot(df, aes(percentB, income)) + 
    geom_point(color = "orange", size = 0.5) + 
    geom_smooth(method = "lm", color ="blue") +
    xlab("Percent of Black People in Tract") +
    ylab("Average Income of Tract") +
    ggtitle("Percent Blacks vs. Average Income (in tract) ")
  # Hispanic
  ggplot(df, aes(percentH, income)) + 
    geom_point(color = "yellow", size = 0.5) + 
    geom_smooth(method = "lm", color ="blue") +
    xlab("Percent of Hispanic People in Tract") +
    ylab("Average Income of Tract") + 
    ggtitle("Percent Hispanics vs. Average Income (in tract) ")
  # Native
  ggplot(df, aes(percentN, income)) + 
    geom_point(color = "green", size = 0.5) + 
    geom_smooth(method = "lm", color ="blue") +
    xlab("Percent of Native People in Tract") +
    ylab("Average Income of Tract") +
    ggtitle("Percent Natives vs. Average Income (in tract) ")
  # Pacific
  ggplot(df, aes(percentP, income)) + 
    geom_point(color = "dodgerblue2", size = 0.5) + 
    geom_smooth(method = "lm", color ="blue") +
    xlab("Percent of Pacific People in Tract") +
    ylab("Average Income of Tract") +
    ggtitle("Percent Pacifics vs. Average Income (in tract) ")
  # White
  ggplot(df, aes(percentW, income)) + 
    geom_point(color = "violet", size = 0.5) + 
    geom_smooth(method = "lm", color ="blue") +
    xlab("Percent of White People in Tract") +
    ylab("Average Income of Tract") + 
    ggtitle("Percent Whites vs. Average Income (in tract) ")
```

## Visualizng Trends through Bar Plots to Predict Incomem
We began by plotting the average income for tracts.
```{r, echo=FALSE}
# Setup variables (considering tracts that have >50% of one race)
  income = clean$Income
  hispanic = which(clean$Hispanic>50) 
  white = which(clean$White>50)
  black = which(clean$Black>50)
  native = which(clean$Native>50)
  asian = which(clean$Asian>50)
  pacific = which(clean$Pacific>50)
  
  ## Income
    #Setup income variables
    mostlywhite = clean$Income[white]
    mostlyblack = clean$Income[black]
    mostlyhispanic = clean$Income[hispanic]
    mostlyasian = clean$Income[asian]
    mostlynative = clean$Income[native]
    mostlypacific = clean$Income[pacific]
    # Get mean incomes
    whiteincome = mean(mostlywhite)
    blackincome = mean(mostlyblack)
    hispanicincome = mean(mostlyhispanic)
    asianincome = mean(mostlyasian)
    nativeincome = mean(mostlynative)
    pacificincome = mean(mostlypacific)
    # Visualize Mean Income by Race
    barplot(c(whiteincome, blackincome, hispanicincome, asianincome, nativeincome, pacificincome), names.arg = c("white", "black", "hispanic", "asian", "native", "pacific"),main = "Income by Predominant Racial Group in Tract", xlab = "Race", ylab = "Income ($)", col = rgb(0,0.3,0.7,0.4))
```

Given that the tracts with predominantly Asian populations (i.e., >50%) had the highest mean income, we tested to see if the income of these tracts were significantly greater than the other tracts.


```{r}
    #Predominantly Asian tracts have the highest average income
    #Is it significantly greater than each other race?
    w = asianincome - whiteincome
    b = asianincome - blackincome
    h = asianincome - hispanicincome
    n = asianincome - nativeincome
    p = asianincome - pacificincome
    
    s.w = var(mostlywhite)/length(mostlywhite)
    s.b = var(mostlyblack)/length(mostlyblack)
    s.h = var(mostlyhispanic)/length(mostlyhispanic)
    s.a = var(mostlyasian)/length(mostlyasian)
    s.n = var(mostlynative)/length(mostlynative)
    s.p = var(mostlypacific)/length(mostlypacific)
    
    t.w = w/sqrt(s.w+s.a)
    t.b = b/sqrt(s.b+s.a)
    t.h = h/sqrt(s.h+s.a)
    t.n = n/sqrt(s.n+s.a)
    t.p = p/sqrt(s.p+s.a)
    
    p.w = 1 - pnorm(t.w, 0, 1)
    p.b = 1 - pnorm(t.b, 0, 1)
    p.h = 1 - pnorm(t.h, 0, 1)
    p.n = 1 - pnorm(t.n, 0, 1)
    p.p = 1 - pnorm(t.p, 0, 1)
    
    p.w; p.b; p.h; p.n; p.p  #All of the differences are significant
```

Ultimately, we found that the differences were clearly statisically significant in all cases.

We also used the percentage of people working in a “professional” industry (management, business, science, and arts) could be used as a predictor of being “wealthy” (median income over $50,000) in Massachusetts using a logistic regression curve. The curve demonstrated a pretty strong positive correlation, and the model gave fairly accurate predictions of the data
```{r, echo=FALSE}
if (!require(stats4)) install.packages("stats4"); library(stats4)
income = clean$Income
  mass = which(clean$State == "Massachusetts") #look at just tracts in Massachusetts
  professional = clean$Professional[mass] #independent variable is the percentage of workers employed in management, business, science, and arts
  hist(professional, breaks = "fd", main = "Distribution of Percentage of Professional Workers in Tract", xlab = "Percent Professional Workers", col = rgb(.4,.6,0,.5))
  
  
  wealthy <- (as.numeric(income[mass]>=50000)); head(wealthy) #threshold for being a "wealthy" county is average income above $50,000
  plot(professional,wealthy,xlab = "Percent Professional Workers", ylab = "Wealthy", main = "Wealthy by Percent Professional")  
  
  MLL<- function(alpha, beta) {
   -sum( log( exp(alpha+beta*professional)/(1+exp(alpha+beta*professional)) )*wealthy
         + log(1/(1+exp(alpha+beta*professional)))*(1-wealthy) )
  }
  results<-mle(MLL, start = list(alpha = 0, beta = 0)) #an initial guess is required
  results@coef
  curve( exp(results@coef[1]+results@coef[2]*x)/ (1+exp(results@coef[1]+results@coef[2]*x)),col = "blue", add=TRUE)
  abline(v=20, col = "red"); abline(v=40, col = "green")
  index <- which(professional >= 19 & professional <= 21); head(index)   #evaluated in a small range of values around 20, because very few tracts were exactly equal to 20 and might not give an accurate estimate
  a = mean(wealthy[index]); a   #21% wealthy tracts 
  abline(h=a, col = "red")  #not bad
  index <- which(professional >= 39 & professional <= 41); head(index)   #evaluated in a small range of values around 40, because very few tracts were exactly equal to 40 and might not give an accurate estimate
  b = mean(wealthy[index]); b   #87% wealthy tracts
  abline(h=b, col = "green")  #pretty good
```


